---
title: "Checking a Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking a Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r}
library(orbitr)
library(dplyr)
library(ggplot2)
```

A simulation always produces *something*. The question is whether the something is physics or numerical artifact, and a plot of the trajectories won't always tell you. This article covers the functions for finding out: `get_energy()`, `get_momentum()`, and `get_angular_momentum()` compute the quantities that real gravity keeps constant, `conserved_quantities()` reports how well a run kept them, `get_orbital_elements()` tells you what orbit a body is actually on, and `continue_simulation()` lets you fix the most common problem — a time step that's too large for part of the run — without re-running everything.

## Energy, Momentum, and Angular Momentum

Newtonian gravity conserves three things exactly: total energy, total linear momentum, and total angular momentum. Three getters compute them at every time step of a run:

```{r}
star_planet <- create_system() |>
  add_body("Star", mass = 1e30) |>
  add_body("Planet", mass = 1e24, x = 1e11, vy = 30000)

sim <- simulate_system(star_planet, time_step = seconds_per_hour * 6,
                       duration = seconds_per_year * 3)

get_energy(sim)
```

`kinetic` is $\sum \tfrac12 m v^2$ over the bodies, `potential` is $-\sum G m_j m_k / r_{jk}$ over the pairs, and `energy` is the total. The two trade off along the orbit — the planet speeds up as it falls inward — while the total stays put:

```{r energy-tradeoff}
energy <- get_energy(sim)

bind_rows(
  energy |> transmute(time, name = "kinetic",   value = kinetic),
  energy |> transmute(time, name = "potential", value = potential),
  energy |> transmute(time, name = "total",     value = energy)
) |>
  ggplot(aes(x = time / seconds_per_day, y = value, color = name)) +
  geom_line() +
  labs(x = "Day", y = "Energy (J)", color = NULL) +
  theme_minimal()
```

`get_momentum()` and `get_angular_momentum()` do the same for $\mathbf{P} = \sum m \mathbf{v}$ and $\mathbf{L} = \sum m\, \mathbf{r} \times \mathbf{v}$:

```{r}
get_momentum(sim)
get_angular_momentum(sim)
```

The total momentum here is not zero — the star was started at rest while the planet moves — so the system's center of mass drifts. That's not an error; it's what `shift_reference_frame(sim, "barycenter")` is for.

## Conserved Quantities

A numerical integrator does not conserve all three exactly, and how badly it fails is a direct measure of how much to trust the run. `conserved_quantities()` joins the three getters and adds each quantity's relative error against its initial value:

```{r}
cq <- conserved_quantities(sim)
cq
```

What should you expect?

- **Momentum and angular momentum** should sit at floating-point rounding, around `1e-15`, for the Velocity Verlet and Euler-Cromer integrators, at *any* time step. Both methods are built from "kicks" (velocity updates) and "drifts" (position updates) that conserve these two quantities exactly. If either one drifts, the problem is in your setup, not the step size.
- **Energy** is conserved only approximately. With Verlet the error oscillates once per orbit and stays inside a band whose width scales as the square of the time step. It should not grow.

Here are all three integrators side by side:

```{r conservation-compare}
runs <- bind_rows(lapply(c("verlet", "euler_cromer", "euler"), function(m) {
  simulate_system(star_planet, time_step = seconds_per_hour * 6,
                  duration = seconds_per_year * 3, method = m) |>
    conserved_quantities() |>
    mutate(method = m)
}))

runs |>
  ggplot(aes(x = time / seconds_per_year, y = energy_error, color = method)) +
  geom_line() +
  facet_wrap(~ method, scales = "free_y", ncol = 1) +
  labs(x = "Years", y = "Relative energy error", color = NULL) +
  theme_minimal()
```

Verlet's error is a bounded wiggle. Euler-Cromer's is a larger bounded wiggle. Euler's climbs without limit — that's the energy being pumped into the orbit that makes it spiral outward. The angular momentum column tells the same story more sharply:

```{r}
runs |>
  group_by(method) |>
  summarise(max_angular_momentum_error = max(angular_momentum_error),
            max_energy_error = max(abs(energy_error)))
```

### Reading the energy plot

Four shapes cover almost everything:

- **A narrow oscillating band** — healthy. Halving the step should shrink it by about four.
- **A wide oscillating band** — the step is too large for the tightest or most eccentric orbit in the system. Halve it and compare.
- **A steady drift** — either `method = "euler"`, or a step so large that even Verlet's guarantees have broken down.
- **A sudden jump** — a close encounter the step couldn't resolve. Everything after the jump is on a different orbit from everything before it.

One thing to know: if you ran with `softening`, the system conserves the *softened* energy. `get_energy()` and `conserved_quantities()` read the softening value the simulation was run with from the output, so this is handled automatically; if you compute energy yourself, use the same softened distance.

## What Orbit Is It Actually On?

`add_body_keplerian()` turns six orbital elements into a position and velocity. `get_orbital_elements()` goes the other way: from the simulated position and velocity of a body relative to a parent, it recovers the semi-major axis, eccentricity, inclination, node, argument of periapsis, and true anomaly at every time step.

```{r}
mars <- create_system() |>
  add_sun() |>
  add_planet("Mars", parent = "Sun", nu = 45) |>
  simulate_system(time_step = seconds_per_day, duration = seconds_per_day * 687)

elements <- get_orbital_elements(mars, "Mars", "Sun")
elements
```

For a two-body system the elements are constants of the motion, so the only thing that changes along the run is `nu`, the position on the orbit. How constant the rest are is another integration check:

```{r}
elements |>
  summarise(a_spread = max(a) - min(a),
            e_spread = max(e) - min(e),
            period_days = mean(period) / seconds_per_day)
```

With a third body the elements are no longer constant — the orbit is being perturbed, and the elements at each instant describe the *osculating* orbit, the ellipse the body would follow from that moment if the perturbation were switched off. Watching them drift is how astronomers describe perturbations. Here is Earth's eccentricity with and without Jupiter:

```{r osculating}
earth_alone <- create_system() |>
  add_sun() |>
  add_planet("Earth", parent = "Sun") |>
  simulate_system(time_step = seconds_per_day, duration = seconds_per_year * 12)

earth_jupiter <- create_system() |>
  add_sun() |>
  add_planet("Earth", parent = "Sun") |>
  add_planet("Jupiter", parent = "Sun", nu = 90) |>
  simulate_system(time_step = seconds_per_day, duration = seconds_per_year * 12)

bind_rows(
  get_orbital_elements(earth_alone, "Earth", "Sun") |> mutate(system = "Sun + Earth"),
  get_orbital_elements(earth_jupiter, "Earth", "Sun") |> mutate(system = "Sun + Earth + Jupiter")
) |>
  ggplot(aes(x = time / seconds_per_year, y = e, color = system)) +
  geom_line() +
  labs(x = "Years", y = "Osculating eccentricity", color = NULL) +
  theme_minimal()
```

The wiggles repeat each time Jupiter laps Earth, and they are real physics, not integration error — the Sun + Earth line is flat.

The elements are also the sharpest test of a time step there is. Energy conservation is necessary but not sufficient: a Verlet run with too coarse a step can keep energy bounded while its orbit slowly *precesses* for no physical reason. If `arg_pe` drifts in a two-body run, that's the integrator, and the drift rate falls by about four when you halve the step.

### A note on the gravitational parameter

By default `get_orbital_elements()` uses $\mu = G M_{\text{parent}}$, the same convention as `add_body_keplerian()`, so elements round-trip exactly. The simulation itself moves both bodies, and the true two-body orbit has $\mu = G(M_{\text{parent}} + m_{\text{body}})$. The difference is negligible for every planet, and about 1% for the Moon; pass `mu = gravitational_constant * (mass_earth + mass_moon)` if you need the exact lunar orbit.

## Continuing a Run

`simulate_system()` uses one fixed time step for the whole run. That's the right design for planets, and the wrong one for a comet: Halley's Comet spends 73 of its 75 years crawling through the outer solar system and a few weeks sprinting around the Sun at sixty times the speed. A step that is fine at aphelion is useless at perihelion.

`continue_simulation()` picks up a run from its last state with whatever step you like, and appends the new rows with `time` continuing from where it left off. So you can take large steps where nothing happens and small ones where everything does:

```{r}
comet <- create_system() |>
  add_sun() |>
  add_body_keplerian("Comet", mass = 1e14, parent = "Sun",
                     a = 5 * distance_earth_sun, e = 0.9, nu = 180)

# Coarse on the way in, fine through perihelion, coarse on the way out
segmented <- comet |>
  simulate_system(time_step = seconds_per_day * 5, duration = seconds_per_year * 4.9) |>
  continue_simulation(time_step = seconds_per_hour * 2, duration = seconds_per_year * 1.4) |>
  continue_simulation(time_step = seconds_per_day * 5, duration = seconds_per_year * 4.9)

n_distinct(segmented$time)
```

Compare with the same orbit run entirely at the coarse step:

```{r continue-compare}
uniform <- simulate_system(comet, time_step = seconds_per_day * 5,
                           duration = seconds_per_year * 11.2)

bind_rows(
  conserved_quantities(segmented) |> mutate(run = "5 day / 2 hour / 5 day"),
  conserved_quantities(uniform)   |> mutate(run = "5 day throughout")
) |>
  ggplot(aes(x = time / seconds_per_year, y = energy_error, color = run)) +
  geom_line() +
  labs(x = "Years", y = "Relative energy error", color = NULL) +
  theme_minimal()
```

The uniform run makes all of its error in the few weeks around perihelion, in one spurious kick, and leaves on a different orbit. The segmented run resolves the passage at a small fraction of the cost of taking two-hour steps for eleven years.

`continue_simulation()` reads the gravitational constant, integrator, and softening from the previous segment, so you only need to say what changes. Each segment is an ordinary Verlet run; the only disturbance is about one step's worth of error at each switch.

### Editing between segments

Because the handoff goes through the output tibble, you can change the state between segments: give a body a velocity kick (a rocket burn), remove one, or add one with `system_from_simulation()`, which rebuilds an `orbit_system` from any snapshot:

```{r}
later <- system_from_simulation(segmented, time = seconds_per_year * 5)
later
```

Modify it with `add_body()` or `remove_body()` like any other system, and simulate again.
