---
title: "Quick Start Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Installation

```r
# install.packages("devtools")
devtools::install_github("daverosenman/orbitr")
```

For 3D interactive plotting, you'll also want:

```r
install.packages("plotly")
```

## Your First Simulation in 30 Seconds

`orbitr` is designed around a simple pipe-friendly workflow: create a system, add bodies, simulate, and plot.

```{r}
library(orbitr)

create_system() |>
  add_body("Earth", mass = mass_earth) |>
  add_body("Moon",  mass = mass_moon, x = distance_earth_moon, vy = speed_moon) |>
  simulate_system(time_step = 3600, duration = 86400 * 28) |>
  plot_orbits()
```

That's it — a 28-day lunar orbit in four lines. Here's what each step does:

1. **`create_system()`** initializes an empty simulation with standard gravitational constant G.
2. **`add_body()`** places a body with a given mass, position, and velocity. All positions are in meters, velocities in m/s. The built-in constants (`mass_earth`, `distance_earth_moon`, etc.) save you from looking anything up.
3. **`simulate_system()`** runs the N-body integration forward in time. `time_step` is how many seconds per integration step, `duration` is the total time to simulate.
4. **`plot_orbits()`** produces a quick 2D trajectory plot using `ggplot2`.

## Customizing the Plot

By default, `plot_orbits()` returns a standard `ggplot` object for planar (2D) simulations and a `plotly` HTML widget for simulations with any 3D motion. (You can also force 3D rendering on planar data with `three_d = TRUE`.) Because the 2D case returns a regular ggplot, you can layer additional geoms, scales, themes, and labels onto it with `+` like any other ggplot.

A common annoyance is that the central body in a two-body system can be invisible: `plot_orbits()` draws each body as a `geom_path()` of its trajectory, and a much more massive body barely moves so its path is too small to see. The Sun in a Sun-Earth simulation is the classic example — it's there, but its loop around the barycenter is well inside the Sun itself. The simplest fix is to drop a marker at the origin:

```{r sun-earth-plot-with-sun, message = FALSE}
sim <- create_system() |>
  add_body("Sun",   mass = mass_sun) |>
  add_body("Earth", mass = mass_earth, x = distance_earth_sun, vy = speed_earth) |>
  simulate_system(time_step = 86400, duration = 86400 * 365)

sim |>
  plot_orbits() +
  ggplot2::geom_point(
    data = data.frame(x = 0, y = 0),
    ggplot2::aes(x = x, y = y),
    color = "gold",
    size = 6
  ) +
  ggplot2::labs(title = "Earth-Sun Orbit")
```

This works because the Sun sits essentially at the origin throughout the simulation. For systems where the central body actually moves a noticeable amount, you'd want to pull its position from the simulation tibble instead of hardcoding `(0, 0)`.

## Adding More Bodies

Since `orbitr` is a full N-body engine, you can add as many bodies as you want. Each one gravitationally interacts with every other. Here's the Sun-Earth-Moon system for a full year:

```{r}
create_system() |>
  add_body("Sun",   mass = mass_sun) |>
  add_body("Earth", mass = mass_earth, x = distance_earth_sun, vy = speed_earth) |>
  add_body("Moon",  mass = mass_moon,
           x = distance_earth_sun + distance_earth_moon,
           vy = speed_earth + speed_moon) |>
  simulate_system(time_step = 3600, duration = 86400 * 365) |>
  shift_reference_frame("Earth") |>
  plot_orbits()
```

Notice `shift_reference_frame("Earth")` — this re-centers everything on Earth so you can see the Moon's orbit instead of having everything overlap at the Sun's scale.

## Changing the Integrator

The default integrator is Velocity Verlet, which conserves energy and keeps orbits stable. You can switch to `"euler_cromer"` for faster (but less accurate) runs, or `"euler"` to see what happens when energy isn't conserved:

```{r}
create_system() |>
  add_body("Star", mass = 1e30) |>
  add_body("Planet", mass = 1e24, x = 1e11, vy = 30000) |>
  simulate_system(time_step = 3600, duration = 86400 * 365, method = "euler") |>
  plot_orbits()
```

You'll see the orbit spiral outward — that's the Euler method artificially pumping energy into the system. Switch back to `method = "verlet"` for a clean closed ellipse.

## The Output is Just a Tibble

`simulate_system()` returns a standard tidy tibble. You can use `dplyr`, `ggplot2`, `plotly`, or any other tool on it:

```{r}
sim <- create_system() |>
  add_body("Earth", mass = mass_earth) |>
  add_body("Moon",  mass = mass_moon, x = distance_earth_moon, vy = speed_moon) |>
  simulate_system(time_step = 3600, duration = 86400 * 28)

sim
```

Each row is one body at one point in time, with columns for position (`x`, `y`, `z`), velocity (`vx`, `vy`, `vz`), mass, body ID, and time.

## Next Steps

- **[The Physics](the-physics.html)** — Understand the math behind the simulation
- **[Examples](examples.html)** — More complex systems including binary stars and the Kepler-16 system
- **[Unstable Orbits](unstable-orbits.html)** — Why most random configurations are chaotic
- **[3D Plotting](plotting-3d.html)** — Interactive 3D visualization with plotly
- **[Custom Visualization](custom-visualization.html)** — Build your own plots with ggplot2 and plotly
- **[Physical Constants](physical-constants.html)** — All built-in masses, distances, and speeds
- **[Roadmap](roadmap.html)** — Features being considered for future versions, plus a place to suggest your own
